---
title: "Driver Genes PAAD"
output: html_notebook
---

# dndscv

## Objective

Identify driver genes in dataset of PAAD (Pancreatic Adenocarcinoma) from TCGA consortium.

## Preparation

``` bash
cd /mnt/data/bioinfo-estadistica-2/lgonzalez/datos_dani/cancer_genomics/data/PAAD

# Copy the BLCA example to personal computer
scp -r ysosa@ken.lavis.unam.mx:/mnt/data/bioinfo-estadistica-2/lgonzalez/datos_dani/cancer_genomics/data/PAAD .
```

## Data loading and exploration

To visualize the data with dndscv, we need an input file of 5 columns, which we can extract from MAF files or VCF files:

1.  Sample ID
2.  Chromosome
3.  Position
4.  Reference
5.  Mutation

The info is in `muts`

```{r}
# Load library
library(dndscv)

# Read the file with 5 columns of bladder cancer
file_mutations = read.table("C:/Users/Dell/Documents/funcional2/PAAD/data_mutations.txt", header=T, sep='\t', stringsAsFactors=F)
head(file_mutations)

#First observations of dataset
muts = file_mutations[ , c("Tumor_Sample_Barcode", "Chromosome", "Start_Position", "Reference_Allele", "Tumor_Seq_Allele2")]
colnames(muts) = c("sampleID", "chr", "pos", "ref", "mut")
head(muts)

clinical_data = read.table("C:/Users/Dell/Documents/funcional2/PAAD/paad_tcga_pan_can_atlas_2018_clinical_data.tsv", header=T, sep='\t', stringsAsFactors=F)
head(clinical_data)
```

Exploring the data base structure

```{r}
# Number of samples
cat("Number of samples: ", length(unique(muts$sampleID)), "\n")

# Number of mutations
cat("Number of mutations: ", nrow(muts), "\n")
```

Mutational burden per sample

```{r}
barplot(sort(table(muts$sampleID)), ylab="Number of mutations", xlab="Donors", las=2, names.arg="", main="Mutational Burden")
```

It's important to exclude the **hypermutators** (more than 500 mutations in the exome) because it can have a negative impact on the statistical process to detect drivers.

```{r}
# Verifica exactamente cuántas muestras caen en cada rango
mut_count = table(muts$sampleID)

sum(mut_count > 1000)   # hipermutadores extremos
sum(mut_count > 500)    # hipermutadores moderados  
sum(mut_count > 200)    # zona gris

# Ver el outlier
sort(mut_count, decreasing = TRUE)[1:5]
```

There's only one hypermutator with over 10,927 mutations.

## Driver detection

### Gene level signals of selection

To detect driver genes, we must remove hypermutators (`max_coding_muts_per_sample=500`) and samples with more than 3 mutation in a given gene (`max_muts_per_gene_per_sample=3`).

The `outmats=T` option generates matrixes with information about the mutation in the trinucleotide context, type of substitution and specific site.

```{r}
dout = dndscv(muts, max_muts_per_gene_per_sample=3,max_coding_muts_per_sample=500, outmats=T)
```

The output lists only unique mutations if the samples are related. Some bases were affected but we'll ignore it because it's a very small number

### Looking at the output

```{r}
names(dout)
```

## Table of significant genes

Analyze genes with significant q-values (combined p-value for the different p-values calculated), which are stored in `sel_cv$qglobal_cv`

```{r}
driver_genes = dout$sel_cv[which(dout$sel_cv$qglobal_cv < 0.1), ]
driver_genes
genes = driver_genes$gene_name
```

There are 7 significative genes: *TP53*, *KRAS*, *CDKN2A.p16INK4a*, *CDKN2A.p14arf,* *SMAD4, RNF43, ARID1A*

```{r}
options(scipen = 999)
dout$sel_cv[which(dout$sel_cv$qglobal_cv < 0.1 & # Verifies it's significant
                    (dout$sel_cv$wnon_cv < 1 | 
                    dout$sel_cv$wspl_cv < 1 | 
                    dout$sel_cv$wind_cv < 1)), ]
```

There are no genes under negative selection.

```{r}
geneci(dout, gene_list = "TP53")
```

## Hotspots

```{r}
for (gene in genes){
  name_gene = dout$annotmuts[which(dout$annotmuts$gene == gene), ]
  cat("Table of gene: ", gene, "\n")
  print(table(name_gene$aachange))
  cat("\n")
}
```

The dots (.) represent impacts in splicing sites, deletions or insertions that change ORF. That's the reason a change in a specific aminoacid isn't shown.

We can do an analysis with more information

```{r}
# Adds column with information about gene, AA change, nucleotide change, position and impact
dout$annotmuts$gene_and_aachange = paste(dout$annotmuts$gene,
dout$annotmuts$aachange, dout$annotmuts$ntchange, dout$annotmuts$pos,
dout$annotmuts$impact, sep=":")

# Counts how many times a observation appears
# Organizes from most to least
# Just first 10 observations
sort(table(dout$annotmuts$gene_and_aachange), decreasing=T)[1:10]
```

We could find more information about the 7 genes in COSMIC. Functions, common mutations and important domains.

## Global signals of selection

To see the global dN/dS ratios for all genes, we can use the following code:

```{r}
print(dout$globaldnds)
```

**Is there evidence of positive or negative selection?**

wspl is \<1, that could mean negative selection, but dndscv interprets the depletion of mutations at splice sites like signs of negative selection.

However, as all the other coefficients and their confidence intervals are \> 1, they may represent positive selection.

But, to estimate the number of missense driver mutations per sample, we have the following information:

```{r}
w = dout$globaldnds[1,2]
n_mis = length(which(dout$annotmuts$impact == "Missense"))
num_samples = length(table(unique(muts$sampleID)))

cat("Selection coefficient of missense mutations: ", w, "\n")
cat("Number of missense mutations in cohort: ", n_mis, "\n")
cat("Number of samples in cohort: ", num_samples)

```

$$
\frac{\omega-1}{\omega}=\frac{1.1436-1}{1.1436}=0.1255
$$

The proportion of missense mutations with positive selection are 12.55%, which represent 326.92 mutations in the cohort. The average per sample is 1.981 mutations.

## Sitednds

Selection at specific sites

```{r}
# Load the the Cancer Gene Census (v81) genes
data("cancergenes_cgc81", package="dndscv")
dout_cancergenes = dndscv(muts, outmats=T, gene_list=known_cancergenes)
```

```{r}
sout = sitednds(dout_cancergenes)
names(sout)
```

```{r}
sout$recursites[which(sout$recursites$qval < 0.1), ]
```

We found some sites with a relevant q-valor in the previously identified genes, some are relevent hotspots.

## Codondnds

Looks for selection at codons.

```{r}
data("refcds_hg19", package = "dndscv")
RefCDS_codon = buildcodon(RefCDS)
codon_dnds = codondnds(dout_cancergenes, RefCDS_codon,
theta_option="conservative", min_recurr=2)

# Output of significant values
codon_dnds$recurcodons[which(codon_dnds$recurcodons$qval<0.1),]
```

# Maftools

Oncoplots or waterfall plots are plot that help visualize frequently affected genes in a cohort, and information like number and type of mutations, annotations of clinical data, etc.

Each column is a sample, and each row is a gene. Shows mutations with color code.

We'll generate them with maftools with our PAAD data.

```{r}
library(maftools)
#Pudimos leerlo directamente porque ya tenía las columnas correctas
PAAD = read.maf(maf = "C:/Users/Dell/Documents/funcional2/PAAD/data_mutations.txt")
```

Ver número de muestras mutadas

```{r}
gen_summary = getGeneSummary(PAAD)
gen_summary
```

```{r}
oncoplot(maf = PAAD)
```

We can see the genes with more mutations.

```{r}
oncoplot(PAAD, top = 9)
```

Oncoplot with genes that we found with dndscv

```{r}
oncoplot(PAAD, genes = c("KRAS", "TP53", "SMAD4", "CDKN2A", "RNF43", "CSMD2", "ARID1A", "TTN", "MUC16"))
```

SMAD4 and CDK2A appear to be mutually exclusive.

```{r}
oncoplot(PAAD, genes = c("CDKN2A", "SMAD4"))
```

## Adding annotations

```{r}
oncoplot(PAAD, genes = c("TP53", "KRAS", "CDKN2A", "SMAD4", "RNF43", "ARID1A"), draw_titv = TRUE)
```

Most mutations are C\>T

## Adding pathways

```{r}
oncoplot(maf = PAAD, 
         pathways = 'smgb', 
         gene_mar = 5, # Changes margin width in gene names
         fontSize = 0.4, # Changes size of gene names
         topPathways = 3) # Displays 3 most affected pathways
```

The more affected pathways are:

-   MAPK signaling -\> KRAS

-   Genome integrity -\> TP53

-   TGFB signaling -\> SMAD4

## Other tools

### Summary plot

```{r}
plotmafSummary(maf = PAAD, addStat = "median", dashboard = TRUE)
```

### Plot for transitions and transversions

```{r}
PAAD_titv = titv(maf = PAAD, plot = FALSE)
plotTiTv(res = PAAD_titv)
```

### Lollipop plot

```{r}
lollipopPlot(maf = PAAD, gene = "TP53", showMutationRate = TRUE, labelPos = c(248, 245, 282, 273))
lollipopPlot(maf = PAAD, gene = "KRAS", showMutationRate = TRUE, labelPos = c(12, 61))
lollipopPlot(maf = PAAD, gene = "CDKN2A", showMutationRate = TRUE, labelPos = c(83, 97, 94))
lollipopPlot(maf = PAAD, gene = "SMAD4", showMutationRate = TRUE)
lollipopPlot(maf = PAAD, gene = "RNF43", showMutationRate = TRUE)
lollipopPlot(maf = PAAD, gene = "ARID1A", showMutationRate = TRUE)
```
